---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

first we make the code to simulate one generation

```{r}
simulate_one_gen_hardy_weinberg <- function(N_aa,N_Aa, N_AA, fitnessaa, fitnessAa, fitnessAA, avgmigrants, mut_rate) {
  # calculate total pop size
  tot_pop <- N_aa+N_Aa+N_AA
  # calculate average fitness
  avg_fit <- weighted.mean(c(fitnessaa, fitnessAa, fitnessAA), c(N_aa,N_Aa, N_AA))
  # calculate deterministic next gen pop size
  next_gen_tot_pop <- tot_pop*avg_fit
  # calculate the allele frequnecies
  a_tot <- 2*N_aa+N_Aa
  A_tot <- 2*N_AA+N_Aa
  a_to_A_mut <- min(rpois(1,a_tot*mut_rate),a_tot)
  A_to_a_mut <- min(rpois(1,A_tot*mut_rate),A_tot)
  p <- (a_tot+A_to_a_mut-a_to_A_mut)/(2*tot_pop)
  q <- 1-p
  # Next generation genotype frequencies; selection is applied here
  det_aa_next <- (p^2)*fitnessaa/avg_fit
  det_Aa_next <- (2*p*q)*fitnessAa/avg_fit
  det_AA_next <- (q^2)*fitnessAA/avg_fit
  
  #migrationrate
  aa_migrants <- rpois(1, avgmigrants)
  # draw offspring according to Poisson distribution
  offsp_aa <- rpois(1, next_gen_tot_pop*det_aa_next) + aa_migrants
  offsp_Aa <- rpois(1, next_gen_tot_pop*det_Aa_next)
  offsp_AA <- rpois(1, next_gen_tot_pop*det_AA_next)
  
  
  return(c(offsp_aa, offsp_Aa,offsp_AA))
}
```

then we use this to simulate a whole population with t_max as maximal generations

```{r}
simulate_pop_HW <- function(N_init_aa, N_init_Aa, N_init_AA, fitnessaa, fitnessAa, fitnessAA, avgmigrants, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_aa, N_init_Aa, N_init_AA)
  # initiate the variables
  pop_new <- c(N_init_aa, N_init_Aa, N_init_AA)
  v <- 0
  success <- 2
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    #define fitness of a and A for each iteration
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_hardy_weinberg(pop_new[1],pop_new[2], pop_new[3], fitnessaa, fitnessAa, fitnessAA,avgmigrants, mut_rate)
   # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new) 
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]+pop_new[3]>=5*(N_init_aa+ N_init_Aa+ N_init_AA)){
      success <- 3
      break
    }else if (pop_new[1]+pop_new[2]+pop_new[3]<=2*avgmigrants){
      success <- 1
      break
    }
  }
  pop_vector <- rbind(pop_vector,success)
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:(t_max+1))[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("aa","Aa","AA")
  # return the result
  return(pop_vector)	
}
```

And then we make a thousand replicates for 0:10 migrates per generation

### This chunk takes very long to run, please just run the one below

```{r eval=FALSE}
N_aa <- 1000
N_Aa <- 0
N_AA <- 0
fitnessaa <- 0.9
fitnessAa <- 0.9
fitnessAA <- 1.1
mut_rate <- 0.01
t_max <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 1000

# initialize data table - where to collect the results
# run the simulation across all chosen parameters
# loop over switch generations
migrants <- 0:10
data_table <- c()
for(avgmigrants in migrants){
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop_HW(N_aa,N_Aa, N_AA, fitnessaa, fitnessAa, fitnessAA, avgmigrants, mut_rate, t_max)
      # determine total population sizes
      total_size <- head(one_run[,1],-1) + head(one_run[,2],-1) + head(one_run[,3],-1)
      #number of generation at which 1.5*n0 is achieved
      number_of_generations <- length(one_run[,1])-1
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==min_size)[1])
      # determine maximal pop size
      max_pop <- max(total_size)
      # determine success
      success <- tail(one_run[,1],1)
      # enter the data into the table
      data_table <- rbind(data_table,c(avgmigrants,min_gen,min_size,number_of_generations,max_pop,success)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
}
```

This outputs a data table which we transformed into: analysis.csv which is faster to load so we can take it from here. The working directory has to be set on the folder where "analysis.csv" is located!! And viridis needs to be installed

```{r}
library("viridis")
data <- read.csv("analysis.csv")
mig0.extinct <- (sum(data$success[data$migration == 0] == "extinction"))
mig0.medium <- (sum(data$success[data$migration == 0] == "medium"))
mig0.max <- (sum(data$success[data$migration == 0] == "reached_max"))
mig1.extinct <- (sum(data$success[data$migration == 1] == "extinction"))
mig1.medium <- (sum(data$success[data$migration == 1] == "medium"))
mig1.max <- (sum(data$success[data$migration == 1] == "reached_max"))
mig2.extinct <- (sum(data$success[data$migration == 2] == "extinction"))
mig2.medium <- (sum(data$success[data$migration ==2] == "medium"))
mig2.max <- (sum(data$success[data$migration == 2] == "reached_max"))
mig3.extinct <- (sum(data$success[data$migration == 3] == "extinction"))
mig3.medium <- (sum(data$success[data$migration == 3] == "medium"))
mig3.max <- (sum(data$success[data$migration == 3] == "reached_max"))
mig4.extinct <- (sum(data$success[data$migration == 4] == "extinction"))
mig4.medium <- (sum(data$success[data$migration == 4] == "medium"))
mig4.max <- (sum(data$success[data$migration == 4] == "reached_max"))
mig5.extinct <- (sum(data$success[data$migration == 5] == "extinction"))
mig5.medium <- (sum(data$success[data$migration == 5] == "medium"))
mig5.max <- (sum(data$success[data$migration == 5] == "reached_max"))
mig6.extinct <- (sum(data$success[data$migration == 6] == "extinction"))
mig6.medium <- (sum(data$success[data$migration == 6] == "medium"))
mig6.max <- (sum(data$success[data$migration == 6] == "reached_max"))
mig7.extinct <- (sum(data$success[data$migration == 7] == "extinction"))
mig7.medium <- (sum(data$success[data$migration == 7] == "medium"))
mig7.max <- (sum(data$success[data$migration == 7] == "reached_max"))
mig8.extinct <- (sum(data$success[data$migration == 8] == "extinction"))
mig8.medium <- (sum(data$success[data$migration == 8] == "medium"))
mig8.max <- (sum(data$success[data$migration == 8] == "reached_max"))
mig9.extinct <- (sum(data$success[data$migration == 9] == "extinction"))
mig9.medium <- (sum(data$success[data$migration == 9] == "medium"))
mig9.max <- (sum(data$success[data$migration == 9] == "reached_max"))
mig10.extinct <- (sum(data$success[data$migration == 10] == "extinction"))
mig10.medium <- (sum(data$success[data$migration == 10] == "medium"))
mig10.max <- (sum(data$success[data$migration == 10] == "reached_max"))
mig0 <- c(mig0.extinct,mig0.medium,mig0.max)/(mig0.extinct+mig0.medium+mig0.max)
mig1 <- c(mig1.extinct,mig1.medium,mig1.max)/(mig1.extinct+mig1.medium+mig1.max)
mig2 <- c(mig2.extinct,mig2.medium,mig2.max)/(mig2.extinct+mig2.medium+mig2.max)
mig3 <- c(mig3.extinct,mig3.medium,mig3.max)/(mig3.extinct+mig3.medium+mig3.max)
mig4 <- c(mig4.extinct,mig4.medium,mig4.max)/(mig4.extinct+mig4.medium+mig4.max)
mig5 <- c(mig5.extinct,mig5.medium,mig5.max)/(mig5.extinct+mig5.medium+mig5.max)
mig6 <- c(mig6.extinct,mig6.medium,mig6.max)/(mig6.extinct+mig6.medium+mig6.max)
mig7 <- c(mig7.extinct,mig7.medium,mig7.max)/(mig7.extinct+mig7.medium+mig7.max)
mig8 <- c(mig8.extinct,mig8.medium,mig8.max)/(mig8.extinct+mig8.medium+mig8.max)
mig9 <- c(mig9.extinct,mig9.medium,mig9.max)/(mig9.extinct+mig9.medium+mig9.max)
mig10 <- c(mig10.extinct,mig10.medium,mig10.max)/(mig10.extinct+mig10.medium+mig10.max)

mig <- matrix(c(mig0,mig1,mig2,mig3,mig4,mig5,mig6,mig7,mig8,mig9,mig10),nrow = 3, ncol = 11)
colors <- magma(3,begin = 0.2,end = 0.8)
barplot(mig,names.arg = 0:10, xlab = "Average Migrants per generation", col = colors, legend = c("extinction","happy ever after","reached max"), ylab = "proportion in %")
```
