---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

At the beginning we are making the function for one generation. It requires the variables N_aa, N_Aa and N_AA as the number of individuals with each genotype, it does further need the input of the fitness of each genotype, the average number of migrants and the mutation rate.

```{r}
setwd("C:/Users/flori/OneDrive/Dokumente/Studium/THEE researchpractical/TEE_Shadowvalley/Final Project")
simulate_one_gen_hardy_weinberg <- function(N_aa,N_Aa, N_AA, fitnessaa, fitnessAa, fitnessAA, avgmigrants, mut_rate) {
  # calculate total pop size
  tot_pop <- N_aa+N_Aa+N_AA
  # calculate average fitness
  avg_fit <- weighted.mean(c(fitnessaa, fitnessAa, fitnessAA), c(N_aa,N_Aa, N_AA))
  # calculate deterministic next gen pop size
  next_gen_tot_pop <- tot_pop*avg_fit
  # calculate the total allele abundance
  a_tot <- 2*N_aa+N_Aa
  A_tot <- 2*N_AA+N_Aa
  # Calculate the number of mutations
  a_to_A_mut <- min(rpois(1,a_tot*mut_rate),a_tot)
  A_to_a_mut <- min(rpois(1,A_tot*mut_rate),A_tot)
  # Calculate the allele frequencies
  p <- (a_tot+A_to_a_mut-a_to_A_mut)/(2*tot_pop)
  q <- 1-p
  # Next generation genotype frequencies; selection is applied here
  det_aa_next <- (p^2)*fitnessaa/avg_fit
  det_Aa_next <- (2*p*q)*fitnessAa/avg_fit
  det_AA_next <- (q^2)*fitnessAA/avg_fit
  # draw the number of migrants
  aa_migrants <- rpois(1, avgmigrants)
  # draw offspring according to Poisson distribution
  offsp_aa <- rpois(1, next_gen_tot_pop*det_aa_next) + aa_migrants
  offsp_Aa <- rpois(1, next_gen_tot_pop*det_Aa_next)
  offsp_AA <- rpois(1, next_gen_tot_pop*det_AA_next)
  # Return the number of individuals of each genotype
  return(c(offsp_aa, offsp_Aa,offsp_AA))
}
```

We use the function for one generation to generate a function which repeats one generation with adapting variables. Adding to the variables and parameters from the function we defined before this function also requires t_max which stops the simulation after t_max generations. If the population reaches 2 \* avgmigrants or 1.5\* N_0.

```{r}
simulate_pop_HW <- function(N_init_aa, N_init_Aa, N_init_AA, fitnessaa, fitnessAa, fitnessAA, avgmigrants, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_aa, N_init_Aa, N_init_AA)
  # initiate the variables
  pop_new <- c(N_init_aa, N_init_Aa, N_init_AA)
  v <- 0
  success <- 2
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_hardy_weinberg(pop_new[1],pop_new[2], pop_new[3], fitnessaa, fitnessAa, fitnessAA,avgmigrants, mut_rate)
   # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new) 
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct = 2* avgmigrants
    if (pop_new[1]+pop_new[2]+pop_new[3]>=1.5*(N_init_aa+ N_init_Aa+ N_init_AA)){
      success <- 3
      break
    }else if (pop_new[1]+pop_new[2]+pop_new[3]<=2*avgmigrants){
      success <- 1
      break
    }
  }
  pop_vector <- rbind(pop_vector,success)
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:(t_max+1))[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("aa","Aa","AA")
  # return the result
  return(pop_vector)	
}
```

We use this function to make 2000 replicates of the each average migration 0,1,2...9,10 and the three modes of inheritance dominant mutation, heterozygote intermediate and recessive mutation. And then we extract the relevant information we want to look at in the data analysis. Specifically this means extracting the following numbers for every run we did:

-   Mean migrants that entered the population

-   Fitness of the heterozygote to see at mode of inheritance

-   Number of generations the simulation has run

-   Minimal population size

-   Generations until minimal population size was reached

-   Final population size

-   Final allele frequency of the wildtype allele

-   The outcome: namely if the population went extinct, reached a maximum or if it did not meet any threshold

This code should not be executed because it takes very long to run!!!
```{r eval=FALSE}
N_aa <- 1000
N_Aa <- 0
N_AA <- 0
fitnessaa <- 0.9
fitnessAa <- 0.9
fitnessAA <- 1.1
mut_rate <- 0.0005
t_max <- 10000

# determine how often to run the simulation for each set of parameters
no_replicates <- 2000
# initialize data table - where to collect the results
# run the simulation across all chosen parameters
# loop over switch generations
migrants <- 0:10
fitness <- c(0.9,1,1.1)
iteration_values <- sort(rep(migrants,no_replicates))
data_table <- data.frame(matrix(NA,nrow = no_replicates*length(migrants)*length(fitness),ncol = 8))
zahl <- 0
set.seed(62016)
for (fitnessAa in fitness){
  for(i in 1:length(iteration_values)){
    one_run <- simulate_pop_HW(N_aa,N_Aa, N_AA, fitnessaa, fitnessAa, fitnessAA, iteration_values[i], mut_rate, t_max)
    # determine total population sizes
    total_size <- head(one_run[,1],-1) + head(one_run[,2],-1) + head(one_run[,3],-1)
    # number of generation at which 1.5*n0 or extinction is achieved
    number_of_generations <- length(one_run[,1])-1
    # determine final pop size
    final_pop_size <- sum(tail(one_run,2)[1,])    
    # determine minimum population size
    min_size <- min(total_size)
    # determine (first) generation at which this population size occurred
    min_gen <- as.numeric(which(total_size==min_size)[1])
    # determine maximal pop size
    max_pop <- max(total_size)
    # determine final allele frequencies
    p_final <- (0.5*one_run[number_of_generations,2]+one_run[number_of_generations,1])/final_pop_size

    # determine success
    success <- tail(one_run[,1],1)
    # enter the data into the table
    data_table[i+zahl,] <- c(iteration_values[i],fitnessAa,number_of_generations,min_size,min_gen,final_pop_size,p_final,success) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
  }
  zahl <- zahl + length(iteration_values)
}

colnames(data_table) <-c("migration_per_gen","fitnessAa","number_of_generations","min_size","min_gen","final_pop_size","p_final","success")
data_table$success <- factor(data_table$success,1:3,c("extinction","medium","reached_max"))
```

This outputs a data table which we transformed into: Data_2000_reps_from_everything.csv which is faster to load so we can take it from here. The working directory has to be set on the folder where "Data_2000_reps_from_everything.csv" is located!! And viridis needs to be installed.

To analyze the data we first make a matrix with the sum of each outcome under each migration rate

```{r}
# Read in the data
ds <- read.csv("Data_2000_reps_from_everything.csv")
# Create a empty matrix with the correct amount of rows and columns
mig <- matrix(NA,nrow = 3, ncol = 11)
# fill in the according numbers
for (i in 0:(ncol(mig)-1)){
  mig[,i+1] <- c(sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==0.9] == "extinction"),
                 sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==0.9] == "medium"),
                 sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==0.9] == "reached_max"))
}
# translate it into %
mig <- mig/20
```

Here we plot the graph with proportions of each outcome at each migration level.

```{r}
library("viridis")
# Defining the colors we want to use
colors <- viridis(3,begin = 0.2,end = 0.8)
# let both graphs be besides eachother
par(mfrow = (c(1,2)))
# Plot all outcomes
barplot(mig, col = colors, names.arg = 0:10, xlab = "mean migrants/ generation", ylab ="proportion in %")
# Plot the reached max seperate
barplot(mig[3,], names.arg = 0:10, col = colors[3],xlab = "mean migrants/ generation", ylab ="proportion in %")
legend("topright", legend = c("Extinction", "Persistence", "Reached Max"), fill = colors, bty = "n", cex = 0.8)
```

now we create a matrix with the outcomes of the other modes of inheritance, namely heterozygote intermediate and dominant mutation. In the data frame th

```{r}
# Create an empty matrix with the right proportions
mig1and1.1 <- matrix(NA,nrow = 3, ncol = 22)
# x is used to count through the whole matrix, not just to column 11
x <- 1
# assign the sum of each outcome to the corresponding field
for (j in c(1,1.1)){
  for (i in 0:(ncol(mig)-1)){
    mig1and1.1[,i+x] <- c(sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==j] == "extinction"),
                      sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==j] == "medium"),
                      sum(ds$success[ds$migration_per_gen == i & ds$fitnessAa ==j] == "reached_max"))
  }
  x <- x+11
}
# Transform the data into %
mig1and1.1 <- mig1and1.1/20
```
Now we plot the heterozygote intermediate outcomes:
```{r}
library("viridis")
# Draw the colors
colors <- viridis(3,begin = 0.2,end = 0.8)
# Settings to make only one plot but leave space for the legend
par(mfrow = c(1, 1), mar = c(5, 4, 4, 6), oma = c(0, 0, 0, 6))
barplot(mig1and1.1[,1:11], col = colors, names.arg = 0:10, xlab = "mean migrants/ generation", ylab ="proportion in %", main = "Heterozygotes with intermediate fitness")
par(xpd = TRUE)  # Allows plotting outside the plot region
legend("right", inset = -0.3, legend = c("Extinction", "Persistence", "Reached Max"), fill = colors, bty = "n", cex = 0.8)
```

... and for the outcomes with the dominant mutation:

```{r}
library("viridis")
# Draw the colors
colors <- viridis(3,begin = 0.2,end = 0.8)
# Settings to make only one plot but leave space for the legend
par(mfrow = c(1, 1), mar = c(5, 4, 4, 6), oma = c(0, 0, 0, 6))
barplot(mig1and1.1[,12:22], col = colors, names.arg = 0:10, xlab = "mean migrants/ generation", ylab ="proportion in %", main = "Dominant mutation")
par(xpd = TRUE)  # Allows plotting outside the plot region
legend("right", inset = -0.3, legend = c("Extinction", "Persistence", "Reached Max"), fill = colors, bty = "n", cex = 0.8)
```
Next we want to analyze the outcomes in a deeper way. First we look at the final population size of the recessive mutation in each migration level:
```{r}
library(ggplot2)
library(dplyr)

# Filter the data for fitnessAa == 0.9
filtered_data <- ds %>%
  filter(fitnessAa == 0.9) %>%
  mutate(final_pop_size = ifelse(final_pop_size == 0,0.1, final_pop_size)) # Transform the 0's to 0.1's to be able to show it on a log scale
# Draw the positions for the threshold lines
line_data <- data.frame(
  migration_per_gen = 1:11,
  threshold = 2 * (0:10)
) %>%
  mutate(threshold = ifelse(threshold == 0, 0.1, threshold))# Transform the 0's to 0.1's to be able to show it on a log scale

# Create a violin plot of final_pop_size against migration_per_gen
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = final_pop_size)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) +  
  scale_y_log10()+
  labs(x = "Mean Migrants per Generation", y = "Final Population Size",
       title = "Distribution of Final Population Size by Migration per Generation\n(fitnessAa = 0.9)") +
  theme_minimal()+
  # Put the lines in place
  geom_segment(data = line_data, 
               aes(x = migration_per_gen-0.25, 
                   xend = migration_per_gen+0.25, 
                   y = threshold, 
                   yend = threshold), 
               color = "red", 
               size = 1)

```
Now we want to take a look at the final allele frequencies
```{r}
library(ggplot2)
library(dplyr)

# Filter the data for fitnessAa == 0.9
filtered_data <- ds %>%
  filter(fitnessAa == 0.9)

# Create a violin plot of final_pop_size against migration_per_gen
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = p_final)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) +  # Violin plot for distribution
  labs(x = "Mean Migrants per Generation", y = "Final wt Allele Frequency",
       title = "Distribution of Final Allele Frequency by Migration per Generation\n Populations with final Population size = 0 not included \n fitnessAa = 0.9") +
  theme_minimal()
```
The heterozygote intermediate scenario also needs to be analyzed for final population size
```{r}
library(ggplot2)
library(dplyr)

# Filter the data for fitnessAa == 1
filtered_data <- ds %>%
  filter(fitnessAa == 1)

# Create a violin plot of final_pop_size against migration_per_gen
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = p_final)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) +  # Violin plot for distribution
  labs(x = "Mean Migrants per Generation", y = "Final wt Allele Frequency",
       title = "Distribution of Final Allele Frequency by Migration per Generation\n Populations with final Population size = 0 not included\n fitnessAa = 1") +
  theme_minimal()



```
Now we take a closer look at the Dominant mutation scenario
```{r}
library(ggplot2)
library(dplyr)

# Filter the data for fitnessAa == 1.1
filtered_data <- ds %>%
  filter(fitnessAa == 1.1)

# Create a violin plot of final_pop_size against migration_per_gen
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = p_final)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) +  # Violin plot for distribution
  labs(x = "Mean Migrants per Generation", y = "Final wt Allele Frequency",
       title = "Distribution of Final Allele Frequency by Migration per Generation\n Populations with final Population size = 0 not included\n fitnessAa = 1.1") +
  theme_minimal()



```
To inspect the hump shaped outcome in reached maximum amount of the initial simulation we ran the simulation another 10'000 times to see if the effect is the same, tis data is stored in the file "eva_data_slay".
```{r}
df3to7 <- read.csv("eva_data_slay.csv")
mig3to7 <- matrix(NA, nrow = 3, ncol = 5)
colnames(mig3to7) <- 3:7
rownames(mig3to7) <- c("extinction","medium","reached_max")
for (i in 1:ncol(mig3to7)){
  mig3to7[,i] <- c(sum(df3to7$success[df3to7$migration_per_gen == i+2] == 1),
               sum(df3to7$success[df3to7$migration_per_gen == i+2] == 2),
               sum(df3to7$success[df3to7$migration_per_gen == i+2] == 3))
}
mig3to7 <- mig3to7/100
par(mfrow = c(1, 2), mar = c(5, 4, 4, 6), oma = c(0, 0, 0, 6))
barplot(mig3to7, col = colors, names.arg = 3:7, xlab = "mean migrants/ generation", ylab ="proportion in %")
barplot(mig3to7[3,], names.arg = 3:7, col = colors[3],xlab = "mean migrants/ generation", ylab ="proportion in %")
par(xpd = TRUE)  # Allows plotting outside the plot region
legend("right", inset = -1, legend = c("Extinction", "Persistence", "Reached Max"), fill = colors, bty = "n", cex = 0.8)

```
To inspect this behaviour further we plot the time it took to reach the maximumon each migration level
```{r}
library(ggplot2)
library(dplyr)

# Filter data where success == 3 (= reached max)
filtered_data <- df3to7 %>%
  filter(success == 3)
summary_data <- filtered_data %>%
  group_by(migration_per_gen) %>%
  summarise(
    count = n()  # Count of values in each category
  )
# Create the plot
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = number_of_generations)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) + 
  labs(x = "Migration per Generation", y = "Number of Generations",
       title = "numbers of generation until maximum was reached") +
  geom_hline(yintercept = 10000, color = "red", linetype = "dashed", size = 1) +
  geom_text(data = summary_data, aes(x = factor(migration_per_gen), y = max(filtered_data$number_of_generations) * 1.05, 
                                     label = paste("n =", count)), 
            color = "black", size = 3) +
  theme_minimal()
```
Because we already have the code, we also analyze the original data set with this plot.
```{r}
library(ggplot2)
library(dplyr)

# Filter data where success == 3
filtered_data <- ds %>%
  filter(success == "reached_max" & fitnessAa == 0.9)
summary_data <- filtered_data %>%
  group_by(migration_per_gen) %>%
  summarise(
    count = n()  # Count of values in each category
  )
# Create the plot
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = number_of_generations)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) + 
  labs(x = "Migration per Generation", y = "Number of Generations",
       title = "numbers of generation until maximum was reached, fitnessAa = 0.9") +
  scale_y_log10()+
  geom_text(data = summary_data, aes(x = factor(migration_per_gen), y = max(filtered_data$number_of_generations) * 1.05, 
                                     label = paste("n =", count)), 
            color = "black", size = 3) +
  theme_minimal()
```
And because we are already at it we also can make it with the other inheritance modes, here with heterozygote intermediate:
```{r}
library(ggplot2)
library(dplyr)

# Filter data where success == 3
filtered_data <- ds %>%
  filter(success == "reached_max" & fitnessAa == 1)
summary_data <- filtered_data %>%
  group_by(migration_per_gen) %>%
  summarise(
    count = n()  # Count of values in each category
  )
# Create the plot
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = number_of_generations)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) + 
  labs(x = "Migration per Generation", y = "Number of Generations",
       title = "numbers of generation until maximum was reached, fitnessAa = 1") +
  scale_y_log10()+
  geom_text(data = summary_data, aes(x = factor(migration_per_gen), y = max(filtered_data$number_of_generations) * 1.05, 
                                     label = paste("n =", count)), 
            color = "black", size = 3) +
  theme_minimal()
```
... and with a dominant mutation
```{r}
library(ggplot2)
library(dplyr)

# Filter data where success == 3
filtered_data <- ds %>%
  filter(success == "reached_max" & fitnessAa == 1.1)
summary_data <- filtered_data %>%
  group_by(migration_per_gen) %>%
  summarise(
    count = n()  # Count of values in each category
  )
# Create the plot
ggplot(filtered_data, aes(x = factor(migration_per_gen), y = number_of_generations)) +
  geom_violin(fill = "skyblue", color = "darkblue", alpha = 0.6) + 
  labs(x = "Migration per Generation", y = "Number of Generations",
       title = "numbers of generation until maximum was reached, fitnessAa = 1.1") +
  scale_y_log10()+
  geom_text(data = summary_data, aes(x = factor(migration_per_gen), y = max(filtered_data$number_of_generations) * 1.05, 
                                     label = paste("n =", count)), 
            color = "black", size = 3) +
  theme_minimal()
```
